### 2022-11-23 abstract.tex
Working compile

    In \cref{ch:design}, we consider approximately optimal Bayesian experimental design
    using an adaptive row sampling algorithm based on DPPs and show that it provides
    good approximations. Through generalizing the proof techniques \todo{Determinant preserving random matrices}
    applied here, in \cref{ch:double_descent} an extension of the DPP-based design is analyzed in
    closed-form for the over-parameterized $n < d$ regime and predicts a double-descent phenomenon in
    linear regression which closely matches empirical experiments.
    In \cref{ch:projections} we isolate the part of the proof involving concentration
    of bilinear forms of matrix resolvents away from the DPP-based design in order to obtain bounds on
    expected projections $X_S^\dag X_S$ when $X_S = S X$ is obtained by sub-Gaussian sketching matrix $S$.

    The second three chapters concern the setting of probabilistic programming, a computational tool
    for specifying probabilistic models and automating inference through MCMC and VI. In particular, we
    study approximating a target density $p(x)$ with approximations $q(x)$ for the purposes of
    importance sampling and variational inference.
    In \cref{ch:lic}, we consider parameterizing $q(x)$ by graph neural networks which condition each
    node on its Markov blanket. This reduces the conditioning sets for a node, resulting in improvements over
    \cite{le2017inference} and run-times which depend on sparsity in the graphical model rather than
    the length of execution traces.
    When the target $p(x)$ is both multivariate and heavy tailed, \cref{ch:ftvi}
    considers the problem of tail anisotropy through both a theoretical and practical perspective.
    We establish that prior fat-tailed estimators \citep{jaini2020tails} are tail isotropic, propose
    an anisotropic approximation (fat-tailed variational inference, FTVI) where
    an anisotropic product base measures is pushed forwards through a bijective neural network,
    and confirm that in practice FTVI improves both density estimation as well as variational inference.
    However, we find in practice FTVI and other tail-adaptive approximations often have trouble optimizing
    the tail parameter. In \cref{ch:hta}, we consider addressing this issue during static analysis of a
    probabilistic program's source code. We define generalized Gamma tail asymptotics for a number of
    elementary distributions and establish how the tail asymptotics are transformed under algebraic transformations
    such as sums and products. This enables a priori computation of tail parameters, which we show
    improves the stability and convergence of a number of inference tasks.

>>>>

    This dissertation considers a range of problems at the intersection of randomized methods
    and statistics. Initially in the first three chapters, we develop novel analyses to recent problems
    in experimental design, double descent, and random projections by using determinantal
    point processes (DPP).
    \Cref{ch:design} consider approximately optimal Bayesian experimental design
    using an adaptive row sampling algorithm based on DPPs and show that it provides
    good approximations. In \cref{ch:double_descent} an extension of the DPP-based design is analyzed in
    closed-form for the over-parameterized $n < d$ regime and predicts a double-descent phenomenon in
    linear regression which closely matches empirical experiments.
    In \cref{ch:projections} we isolate the part of the proof involving concentration
    of bilinear forms of matrix resolvents away from the DPP-based design in order to obtain bounds on
    expected projections $X_S^\dag X_S$ when $X_S = S X$ is obtained by sub-Gaussian sketching matrix $S$.

    In the later chapters we focus on probabilistic programming and developing theory and tools
    to automate statistical inference using randomized algorithms based on
    Monte Carlo Markoc chain (MCMC) and variational inference (VI).
    \Cref{ch:lic} considers parameterizing $q(x)$ by graph neural networks which condition each
    node on its Markov blanket. This reduces the conditioning sets for a node, resulting in improvements over
    \cite{le2017inference} and run-times which depend on sparsity in the graphical model rather than
    the length of execution traces.
    When the target $p(x)$ is both multivariate and heavy tailed, \cref{ch:ftvi}
    considers the problem of tail anisotropy through both a theoretical and practical perspective.
    However, we find in practice FTVI and other tail-adaptive approximations often have trouble optimizing
    the tail parameter. In \cref{ch:hta}, we consider addressing this issue during static analysis of a
    probabilistic program's source code.
